HO unification from object language to meta language

### 2024-03-23 paper/main.tex
section4

\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$. We assume that the unification of our meta
language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{\Fo{} and \Ho language}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

@@

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it to enable the instantiation of a single
solution to different scopes.
On the contrary a solution to a \Fo variable is a plain term.

@@

\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\end{invariant}

\noindent The arity of a variable in \Ho (a \elpiIn{hvariable} is stored in
the mapping. In particular \elpiIn{m-alloc}
bla bla explain. Multiple mappings for the same \elpiIn{fvariable}
are handled in section \ref{sec:eta}.

@@

\subsection{Notations}

When we write \Ho terms outside code blocks we prefer to use the
usual and more compact mathematical notation. However we need to
distinguish between the ``application'' of a unification variable
to its scope and the application of a term (possibly a unification
variable) to a list of arguments. We write the scope in subscript
and we use juxtaposition for application. Here a few examples:
\vspace{5pt}

\begin{tabular}{ll}
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
\end{tabular}
\vspace{5pt}

\noindent
When detailing example we prefer to write links in mathematical notation,
as equations between terms in a context. The equality sign is subscripted with
kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
\begin{elpicode}
abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
\end{elpicode}

@@

here we give the functions/signatures to express the properties 3-5 in the problem
statement

Together with the description of the terms of the language, we need some
auxiliary functions to perform operations like term equality, unification,
dereferencing. There predicates are supposed to be implemented in the OL and the
ML following respectively their specification. In the following few paragraphs
we give a brief proposition of them.

\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}

Since in our encoding we explicitly carry a substitution we need to
define the operation that applies it to a term. Its proposition
takes care to normalize (flatten) applications, for example
it turns
\elpiIn{(app [app [con"f",con"a"],con"b"])}
into
\elpiIn{(app [con"f",con"a",con"b"])}.

dereference variables of the two languages. This is particularly useful to check
if two terms in the OL (resp. in the ML) are equal. The constructors
representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
make the dereferencing procedure to recurse over the their subterms. The code
below display this behavior for the dereferencing performed by the OL:

@@

\noindent
TODO explain napp.
We use the cut operator to keep the code compact. It is possible to rewrite
the rule for application traversal so that it is mutually exclusive with the
first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.

The corresponding code for \Ho is similar, we only show the last two rules that
differ in a substantial way:

@@

\noindent
TODO: no need to napp, see the beta section.
Note that when the substitution \elpiIn{S} maps a unification variable
\elpiIn{N} to an assignment \elpiIn{F} we ....

TODO: invariant: variables in subst are always fully applied, and length of scope is the
arity of the HO variable.

Important!!!
A different reasoning is to be addressed to the variables of the ML. Firstly, a
meta variable cannot appear in the \elpiIn{app} node as the first element of the
list, we will explain why in \cref{sec:compilation}

\paragraph{Term equality: \Eo vs. \Ee} We can test if two terms are equal
following the equational theory of the language being considered. In
\cref{code:eq-pred} we provide an implementation of the \Eo predicate. The first
four rules check if the two terms are equal regarding the structure of the
current node, that is, two terms are equal if they have same head and if
recursively each subterm is two by two equal. Moreover, since the theory of the
OL accepts $\eta$- and $\beta$-equivalence, then we implement the corresponding
rules.

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
  app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
  con C ~\Ee~fcon C.
  uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
\end{elpicode}

\noindent
The equality relation for the ML, accepts $\eta\beta$-equivalence between terms
of the ML. Recall that \elpiIn{abs x\ f x}, is a valid $\eta$ expansion of the
function \elpiIn{f} and that \elpiIn{lam x\ app[f, x]} is not that equivalent to
\elpiIn{f} at meta level. However, since we are interested in using the
unification procure of the ML, by \cref{prop:correct-ml}, we can use the
\Ue relation to test, when needed if two terms are equal in the ML.

\paragraph{Term unification: \Uo vs. \Ue} 
The last but not least important relation we should take care of before
presenting our full algorithm aiming to unify terms of the OL in the ML and
provide the substitution produced in the ML to the OL, is term unification. This
procedure is a more powerful version of the equal predicate, since unification
checks if two terms can be equal be assigning unification variables. In our
representation, variable assignment (or refinement) is performed by modifying
the corresponding substitution mapping. We will not give an implementation of
\Uo, since we are giving an implementation of it using
our algorithm, see \cref{XX}.

@@

On the other hand, unification in the ML needs to be defined. In
\cref{code:ue-pred}, we give an implementation of \Ue but
that is actually what our meta language provides as a builtin.

This predicate has four
arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
\subseteq \rho_2$.
% The first three rules unify terms with same rigid heads, and
% call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
% assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
% unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
% terms are unification variables, we test that their arguments are in the pattern
% fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
% pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
% $w$ and return the new mapping $\rho_2$ containing all the new variable
% assignment. Finally, if only one of the two terms is an unification variable
% $v$, after having verified that $v$ does not occur in the other term $t$, we
% bind $v$ to $t$ and return the new substitution mapping.

>>>>

\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:problem-statement}

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$. We assume that the unification of our meta
language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

@@

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

@@

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it to enable the instantiation of a single
solution to different scopes. This is obtained via the \elpiIn{inctx}
container, and in particular via its \elpiIn{abs} binding constructor. 
On the contrary a solution to a \Fo variable is a plain term.

@@

\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\end{invariant}

\noindent
The compiler establishes a mapping between variables of the two languages.
In order to preserve invariant \ref{inv:uvaarity} we store the
arity of each \elpiIn{hvariable} in the mapping and we reuse an existing
mapping only if the arity matches.

\noindent
When a single \elpiIn{fvariable} occurs multiple times with different numbers
of arguments the compiler generates multiple mappings for it, on a first
approximation, and then makes the mapping bijective by introducing
\linketa; this detail is discussed in section \ref{sec:eta}.

As we mentioned in section~\ref{sec:nutshell} the compiler
replaces terms in \maybebeta and \maybebeta with fresh variables
linked to the problematic terms. Each class of problematic terms
has a dedicated link.

@@

\begin{invariant}[Link left hand side] The left hand side of a new link
  is a flexible term.
\end{invariant}

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container.

\subsection{Notational conventions}

When we write \Ho terms outside code blocks we follow the
usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
However we need to
distinguish between the ``application'' of a unification variable
to its scope and the application of a term to a list of arguments.
We write the scope of unification variables in subscript
while we use juxtaposition for regular application.
Here a few examples:
\vspace{5pt}

\begin{tabular}{ll}
  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
\end{tabular}
\vspace{5pt}

\noindent
When detailing examples we write links as equations between terms under a context.
The equality sign is subscripted with
kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
\begin{elpicode}
abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
\end{elpicode}

@@

In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
we need to equip \Fo and \Ho with term equality,
substitution application and unification.

\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
over ground terms to the full languages by adding the reflexivity of
unification variables (a variable is equal to itself).

The first four rules are common to both equalities and correspond to $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
  con C ~\Ee~fcon C.
  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
\end{elpicode}

\noindent
The main point in showing these equality tests is to remark how weaker \Ee is,
and to identify the four rules that need special treatment in
the implementation of \Ue.

For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
position in \elpiIn{T} whenever the list \elpiIn{A} provides a corresponding
argument.

\input{code/beta_fo}

\noindent
The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
variables). Elpi provides it as a builtin, but one could implement it by
systematically loading the hypothetical rule \elpiIn{name x} every time
a nominal constant is postulated via \elpiIn{pi x\ }.

\paragraph{Substitution application: $\rho s$ and $\sigma t$}

Applying the substitution corresponds to dereferencing a term with respect to
the memory. To ease the comparison we split \Fo dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, ans has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp},
in charge of ``flattening'' \elpiIn{fapp} nodes, has no corresponding
operation in \Ho. The reasons for this asymmetry is that an \elpiIn{fapp}
node with a flexible head is always mapped to a \elpiIn{uva} (as per
sections \ref{sec:compilation,sec:beta}), preventing nested applications to
materialize.

@@

\noindent
TODO: about the cut\marginpar{should we rename forall2 into map}

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification variable and the
number of abstractions in its assignment have to match.

\paragraph{Term unification: \Uo vs. \Ue}

In this paper we assume to have an implementation of \Ue that satisfies
properties \ref{prop:correct-ml,prop:complete-ml}. Although we provide an
implementation in the appendix (that we used for testing purposes) we only
describe its signature. Elpi is expected to provide this brick, as well as
any other implementation of $\lambda$Prolog.

@@

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\marginpar{explain better}

% The first three rules unify terms with same rigid heads, and
% call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
% assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
% unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
% terms are unification variables, we test that their arguments are in the pattern
% fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
% pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
% $w$ and return the new mapping $\rho_2$ containing all the new variable
% assignment. Finally, if only one of the two terms is an unification variable
% $v$, after having verified that $v$ does not occur in the other term $t$, we
% bind $v$ to $t$ and return the new substitution mapping.
